Fluid-mechanical erosion creates anisotropic porous media

### 2022-06-02 docs/PhysicsPaper/StokesErosion.tex
NM

%-----------------------------------------------------------------------------------------------%
\subsection{Permeability}

@@

%-----------------------------------------------------------------------------------------------%
\subsection{Total drag}

@@

Projecting \eqref{divthm} onto $\ex$, using \eqref{eqn:drag}, and rearranging gives the exact relationship
\begin{equation}
\label{drag_press}
\FD \cdot \ex = \int_{-1}^{1} \left( -p(x,y) + 2 \mu u_x(x,y) \right) \big|_{x=-x_0}^{x=x_0} dy + 
\int_{-x_0}^{x_0} \mu u_y(x,y) \big|_{y=-1}^{y=1} dx
\end{equation}
We will now make some simplifying assumptions. First, because the imposed flow profile is uniform $\UU = (U,0)$, slip is permitted along the top and bottom boundaries $y=\pm 1$ and so the viscous stress will generally be much smaller than on the no-slip erodable boundaries. We will therefore drop the contribution from the second integral above. Second, if $x_0 > 1$ is chosen a sufficient distance from the erodable bodies, the flow profile will approximately match the uniform profile, implying that the term involving $u_x(\pm x_0, y)$ can be dropped. In addition, the pressure $p(\mp x_0, y)$ will approximately match the upstream and downstream values respectively. With these assumptions, \eqref{drag_press} simplifies to the approximate form
\begin{equation}
\FD \cdot \ex \approx 2 (\pup - \pdn)
\end{equation}
Combining with \eqref{eqn:k11} yields an approximate relationship between the horizontal components of permeability and drag 
\begin{equation}
\label{}
\frac{1}{k_{11}} \approx \frac{1}{4 U}  \, \FD \cdot \ex
\end{equation}

USEFUL FOR BENCHMARK

@@

The anisotropy of permeability is
\begin{equation}
\anis = k_1 / k_2
\end{equation}

There are two possible explanations for how erosion creates anisotropy in the microstructure. First, the shear stresses could carve each individual body into a more slender form, thus creating anisotropy at the level of individual grains. Second, by removing material in a non-uniform way, could preferentially disintegrate certain bodies before others, leaving an overall configuration that exhibits anisotropy on the large scale. We will refer to these two possible mechanisms as {\em shape} anisotropy and {\em configurational} anisotropy respectively. As an extreme example, a tightly-packed horizontal row of circular bodies would exhibit configurational anisotropy but no shape anisotropy. Meanwhile, an array of highly eccentric ellipses, all oriented horizontally but positioned randomly, would exhibit high shape anisotropy and low configurational anisotropy.

>>>>

%-------------------------------------------------------------------%
\subsection{Longitudinal permeability}
\label{LongPerm}

@@

%-------------------------------------------------------------------%
\subsection{Relationship to drag}

@@

Projecting \eqref{divthm} onto $\ex$, using \eqref{eqn:drag}, and rearranging gives the exact relationship
\begin{equation}
\label{drag_press}
\FD \cdot \ex = \int_{-1}^{1} \left( -p(x,y) + 2 \mu u_x(x,y) \right) \big|_{x=-x_0}^{x=x_0} dy + 
\int_{-x_0}^{x_0} \mu u_y(x,y) \big|_{y=-1}^{y=1} dx
\end{equation}
We will now make some simplifying assumptions. First, because the imposed flow profile is uniform $\UU = (U,0)$, slip is permitted along the top and bottom boundaries $y=\pm 1$ and so the viscous stress will generally be much smaller than on the no-slip erodable boundaries. We will therefore drop the contribution from the second integral above. Second, if $x_0 > 1$ is chosen a sufficient distance from the erodable bodies, the flow profile will approximately match the uniform profile, implying that the term involving $u_x(\pm x_0, y)$ can be dropped. In addition, the pressure $p(\mp x_0, y)$ will approximately match the upstream and downstream values respectively. With these assumptions, \eqref{drag_press} simplifies to the approximate form
\begin{equation}
\FD \cdot \ex \approx 2 (\pup - \pdn)
\end{equation}
Combining with \eqref{eqn:k11} yields a formula relating the drag and permeability in the longitudinal direction
\begin{equation}
\label{perm_drag}
\frac{1}{k_{11}} \approx \frac{1}{4 U}  \, \FD \cdot \ex
\end{equation}
This formula serves as a valuable benchmark linking the microscopic (Stokes) and macroscopic (Darcy) perspectives. Since the drag is a quantity that can be computed with high accuracy with our Stokes solvers, verifying the relationship \eqref{perm_drag}, as will be done in Fig.~\ref{fig2}, validates a number of approximations made in the coarse-grained approach, such as (1) the homogenization of the flow field implicit in Darcy's law \eqref{eqn:Darcy}; (2) the calculation of the permeability from measurements of the flow rate and upstream/downstream pressures through \eqref{eqn:k11}.

%-------------------------------------------------------------------%
\subsection{Transverse permeability and anisotropy}

Now consider measuring the transverse permeability $k_{22}$ using Darcy's law \eqref{eqn:Darcy}. Taking the vertical component of \eqref{eqn:Darcy} yields
\begin{equation}
\label{q1}
q_2(x,y) = -k_{22} \pderiv{p}{y}(x,y)
\end{equation}
but this form is not useful if no vertical pressure gradient is imposed, as is the case in our erosion simulations. It is important to recognize that the permeability is a property of the {\em medium}, not the imposed flow. Hence, for a frozen configuration of bodies, it is permissible to alter the imposed flow for the purpose of measuring $k_{22}$ (this altered flow is completely separate from the simulation of the erosion process that generates the configurations). Thus, instead of a horizontal flow in the far-field, we seek to impose a vertical one $\UU = (0,U)$.
If this were to be done directly, the outer geometry would need to be rotated by 90 degrees about the configuration of bodies inside (which does not rotate). In practice, it is simpler to keep the outer geometry fixed and instead rotate the inner configuration of bodies, then simply apply the method from Section \ref{LongPerm} to measure $k_{22}$.

With both the longitudinal and transverse components of permeability calculated, we define the anisotropy of the medium as the ratio
\begin{equation}
\anis = k_{11} / k_{22}
\end{equation}
Note that the random configuration of circles used to initialize the erosion simulations will have an anisotropy nearly equal to one, and as the configuration erodes, the configuration would be expected to permit flow in the longitudinal direction more easily than in the transverse direction, yielding $\anis > 1$.

There are two mechanistic explanations for how erosion can create this anisotropy. First, the shear stresses could carve each individual body into a more slender form, thus creating anisotropy at the level of individual grains. Second, by removing material in a non-uniform way, could preferentially disintegrate certain bodies before others, leaving an overall configuration that exhibits anisotropy on the large scale, for example one with horizontally aligned channels. We will refer to these two possible mechanisms as {\em shape} anisotropy and {\em configurational} anisotropy respectively. As an extreme example, a tightly-packed horizontal row of circular bodies would exhibit configurational anisotropy but no shape anisotropy. Meanwhile, an array of highly eccentric ellipses, all oriented horizontally but positioned randomly, would exhibit high shape anisotropy and low configurational anisotropy.

@@
